---
title: "Simulation from Models with Selection"
author: 
date: 
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions for Simulation

$\texttt{rdir.tree}$ produces a random realization of the ancestral selection graph as described in section 4.8 of Hein, Schierup, and Wiuf 2005. The inputs and outputs of this function are detailed below.

Inputs:
\begin{itemize}
  \item $N$: the desired sample size. Should be an integer greater than 1.
  \item $\sigma$: the scaled selection coefficient. Should be a nonnegative real number. Larger values correspond to a stronger selective advantage. Zero corresponds to a neutral model (no selective advantage).
\end{itemize}

Output: A structure containing the information necessary to construct a reticulating tree, as in the left side of Hein et al. Figure 4.21, including branch lengths. This structure is a matrix whose rows represent branching or coalescing events in the dual process. Rows are structured as follows:
\begin{itemize}
  \item The first entry in each row is the time since the event occurred, the "dual" time. Events are listed in the order they occurred. So the time of the first event is the time at which the last two lineages in the dual process coalesced into the lineage of the ultimate ancestor. Time zero represents the present, the start of the dual process.
  \item The second entry in each row is either 1 or 0, where 1 indicates a coalescing event, and 0 represents a branching event.
  \item If the event is a coalescing event, the 4th and 5th entries of the row are the numbers of the two lineages coalescing, and the 3rd entry is the number of the new lineage. (The process begins with $N$ lineages numbered 1 through $N$.)
  \item If the event is a branching event, the 4th and 5th entries of the row are the numbers of the new lineages, the 4th being the continuing branch and the 5th being the incoming branch. The 3rd entry is the number of the lineage that branches.
\end{itemize}
```{r}
rdir.tree = function(N, sigma)
{
  time = 0
  active_lines = 1:N
  next_line = N + 1
  tree = c()
  
  # We generate a new event with each iteration of this loop.
  while(length(active_lines) > 1)
  {
    j = length(active_lines)
    i.coal = j * (j - 1) / 2
    i.branch = sigma * j / 2
    
    # Type of the event: coalescing or branching
    coal = runif(1) < (i.coal / (i.coal + i.branch))
    
    wait.time = rexp(1, rate = i.coal + i.branch)
    time = time + wait.time
    event = c(time, coal)
    
    if(coal)
    {
      lines = sample(1:j, 2)  # sample 2 active lines without replacement
      event = c(event, next_line, active_lines[lines])
      active_lines = c(active_lines[-lines], next_line)
      next_line = next_line + 1
    }
    else
    {
      line = sample(1:j, 1)  # sample 1 active line
      event = c(event, active_lines[line], next_line, next_line + 1)
      active_lines = c(active_lines[-line], next_line, next_line + 1)
      next_line = next_line + 2
    }
    
    tree = rbind(event, tree)  # add the event to the tree
  }
  rownames(tree) = NULL
  colnames(tree) = c("event time", "event type", "[,3]", "[,4]", "[,5]")
  return(tree)
}
```

Example of use:
```{r}
set.seed(43)
tree = rdir.tree(4, 1)
tree
```


TODO: Add a function to print reticulating trees.

TODO: Add documentation for function to generate mutations along a reticulating tree.
Generates mutations according to Jukes Cantor model.
```{r}
rdir.mut = function(tree, u)
{
  mut_tree = c()
  tree = rbind(tree, c(0, -1, 0, 0, 0))  # add a line for the present
  time = tree[1, 1]  # time of the ultimate ancestor
  active_lines = c(tree[1, 3])  # line number of the ultimate ancestor
  for(event in 1:(length(tree[, 1])))
  {
    next_mut = NULL
    while(time > tree[event, 1])
    {
      mut_tree = rbind(mut_tree, next_mut)
      time = time - rexp(1, rate = u * length(active_lines))
      next_mut = c(time, 2, sample(active_lines, 1), 0, 0)
    }
    
    mut_tree = rbind(mut_tree, tree[event, ])
    time = tree[event, 1]
    
    # now just need to update active_lines
    if(tree[event, 2] == 0)
    {
      active_lines = c(active_lines[!(active_lines %in% tree[event, 4:5])],
                       tree[event, 3])
    }
    else if(tree[event, 2] == 1)
    {
      active_lines = c(active_lines[active_lines != tree[event, 3]],
                       tree[event, 4:5])
    }
    # no need to make an update at the sampling event
  }
  rownames(mut_tree) = NULL
  return(mut_tree)
}
```

Example of use:
```{r}
set.seed(2021)
mut_tree = rdir.mut(tree, 1)
mut_tree
```


TODO: Add documentation for function that produces the sample given the type of the ultimate ancestor and the graph with mutations.
```{r}
rdir.sample = function(mut_tree, ua)
{
  # active lineages and their types
  active_lines = matrix(c(mut_tree[1, 3], ua), ncol = 2)
  
  for(event in 1:(length(mut_tree[, 1])))
  {
    if(mut_tree[event, 2] == 1)
    {
      index = which(active_lines[, 1] == mut_tree[event, 3])
      active_lines = rbind(active_lines[-index, ],
                           c(mut_tree[event, 4], active_lines[index, 2]),
                           c(mut_tree[event, 5], active_lines[index, 2]))
    }
    else if(mut_tree[event, 2] == 2)
    {
      index = which(active_lines[, 1] == mut_tree[event, 3])
      active_lines[index, 2] = !active_lines[index, 2]
    }
    else if(mut_tree[event, 2] == 0)
    {
      indices = which(active_lines[, 1] %in% mut_tree[event, 4:5])
      # the type with advantage is 1
      type = active_lines[indices[1], 2] | active_lines[indices[2], 2]
      active_lines = rbind(active_lines[-indices, ],
                           c(mut_tree[event, 3], type))
    }
  }
  return(active_lines[, 2][order(active_lines[, 1])])
}
```

Example of use:
```{r}
sample = rdir.sample(mut_tree, 1)
sample
```

TODO: Add documentation for function that computes the likelihood of a sample given the graph, the mutation rate, and the type of the ultimate ancestor. Note that to get a complete likelihood, not given the common ancestor, we can integrate this function over the stationary distribution of the type of the common ancestor. This amounts to summing $p$ terms, where $p$ is the number of types.
```{r}
# under our simple model, we can compute the likelihood of having an even
# (or odd) number of mutations along a particular lineage given the mutation
# rate and the amount of time the lineage existed
lineage.ll.helper = function(even, u, time)
{
  ll = (1 + exp(-2 * u * time)) / 2
  if(even)
  {
    return(ll)
  }
  return(1 - ll)
}

rdir.llh.summand = function(y, z, m, tree, u, ua)
{
  tree = rbind(tree, c(0, -1, 0, 0, 0))
  z = intToBits(z)[1:m]  # types of latent variables
  z = c(y, z, ua)  # types of all variables
  parent_types = rep(-1, length(z) - 1)
  line_lengths = rep(0, length(z))
  # fill those in
  
  time = tree[1, 1]
  active_lines = c(length(z))
  for(event in 1:length(tree[, 1]))
  {
    new_time = tree[event, 1]
    line_lengths[active_lines] = line_lengths[active_lines] + (time - new_time)
    time = new_time
    if(tree[event, 2] == 1)
    {
      parent_types[tree[event, 4:5]] = z[tree[event, 3]]
      active_lines = c(active_lines[active_lines != tree[event, 3]],
                       tree[event, 4:5])
    }
    else if(tree[event, 2] == 0)
    {
      type = z[tree[event, 4]] | z[tree[event, 5]]
      parent_types[tree[event, 3]] = type
      active_lines = c(active_lines[!(active_lines %in% tree[event, 4:5])],
                       tree[event, 3])
    }
  }
  
  helper = function(i)
  {
    even = (parent_types[i] == z[i])
    return(lineage.ll.helper(even, u, line_lengths[i]))
  }
  return(prod(sapply(1:(length(z) - 1), helper)))
}

rdir.llh = function(y, tree, u, ua)
{
  m = tree[1, 3] - 1 - length(y)  # number of internal nodes to sum over
  helper = function(z)
  {
    rdir.llh.summand(y, z, m, tree, u, ua)
  }
  return(sum(sapply(0:(2^m - 1), helper)))
}
```

Example of use:
```{r}
rdir.llh(sample, tree, 1, 1)
```


TODO: Add documentation for function that validates likelihood calculation through simulation. Last argument is number of iterations.
```{r}
rdir.emp.llh = function(y, tree, u, ua, B)
{
  counter = 0
  for(i in 1:B)
  {
    m_tree = rdir.mut(tree, u)
    samp = rdir.sample(m_tree, ua)
    if(identical(samp, y))
    {
      counter = counter + 1
    }
  }
  return(counter / B)
}
```

Example of use:
```{r}
set.seed(2022)
rdir.emp.llh(sample, tree, 1, 1, 10^4)
```
Does this agree with the likelihood calculation? Which one runs faster?
Yes, they agree! The first is much faster.

Here's another example from start to finish. Change the seed to see different examples. Seed 20 leads to a complicated graph, and our method still works even in this more computationally intense scenario. In this case there are 17 internal nodes. Due to computational limitations, our function will not work if the number of internal nodes exceeds 31. For an example of this, leave the seed at 20, and set the value of sigma to 2. This limitation could probably be improved at the expense of having less readable code and spending a significant amount of time rewriting code.
```{r}
set.seed(20)
N = 4
sigma = 1
u = 1
ua = 1
B = 10^4
tree = rdir.tree(N, sigma)
tree
mut_tree = rdir.mut(tree, u)
mut_tree
sample = rdir.sample(mut_tree, ua)
sample
rdir.llh(sample, tree, u, ua)
rdir.emp.llh(sample, tree, u, ua, B)
```




TODO: Modify function for case of more than 2 types.



